Mathematical Gene Network Evolution

### 2017-06-07 attic/kalman_walk_edit.tex
added to attic

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

\paragraph{To do, soon:}

\begin{enumerate}

  \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction. Still needs a lot of refinement.} \ref{sec:intro}

    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
        \begin{itemize}
            \item \emph{(peter)} Work out the math for how to do this.
            \item \emph{(josh)} Find real parameters to put into this. 
              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
        \end{itemize}

      \item \emph{(josh)} Write out simple examples in biological words. \ref{sec:examples}

    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.

    \item \emph{(josh)} Clarify our assumptions around $x(0)$.

    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.

    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.

    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change. \jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}

    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
        \begin{itemize}
            \item \emph{(josh)} Work out parameterization.
            \item \emph{(peter)} Describe Brownian motion.
        \end{itemize}

    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
        OR consider Josh's parameteriziation like this;
        OR just say we do Brownian motion on the manifold (more or less implicit characterization).

    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
        Ask questions of stationary distribution: robustness, connectivity, xxx?

    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
      First try on simple concrete examples. \jss{see \ref{sec:del}}

    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
        Do we need to allow deleterious variation to get deletion?

    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.

    \item Simulate.

\end{enumerate}

@@

%%% Questions:

% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?

\title{The Evolution of Phenotypically Invariant Genetic Networks}
\author[1]{Joshua S. Schiffman}
\author[1,2]{Peter L. Ralph}
\affil[1]{University of Southern California, Los Angeles, California}
\affil[2]{University of Oregon, Eugene, Oregon}
\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
\maketitle

\begin{abstract}
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
\end{abstract}

\section{Background} \label{sec:intro}

@@

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

@@

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).

@@

Organisms often have to respond to their environments,
and the way this happens is that external input triggers a cascade of molecular signals
whose eventual result is a reaction,
whose approprateness for the situation determines fitness.
\plr{A similar situation occurs in development (...?).}
However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
especially given a number of possible molecular intermediaries,
there may be a large number of ways to construct such a molecular signalling system
that has precisely the \emph{same} input-output relationships,
and hence the same function.
Therefore, any mutation that changes one such system to another, equivalent version
will be neutral,
and may well become polymorphic in a population if there are no ill effects in heterozygotes.
Certainly, mutations that move a system away from optimal functioning will also appear,
but in very large populations will not rise to high frequency.
\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}

@@

\begin{example}[Example: Oscillator.] \label{ex:osc}
Motivate the system:
\begin{align*}
    A &= \left[\begin{matrix} 
        0 & 1 \\ 
       -1 & 0 
    \end{matrix}\right]
\end{align*}

    \begin{figure}
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$x_{1}$};
                \node[dashed] (B) at (3,0) {$x_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every node/.style={fill=white,circle},
                          every edge/.style={draw=black, thick}]
                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
            \end{scope}
            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
            \end{scope}

            \end{tikzpicture}
        \end{center}
        \caption{
            Diagram of Example \ref{ex:osc} in the text.  
            \plr{explain what arrows mean if nec}
            \label{fig:oscillator_diagram}}
    \end{figure}

as a special case of the below.

@@

\begin{example}[Hyrbid Inviability]
  Let, \begin{align*}
    A(0) =\left[\begin{matrix}
      0 & 1 \\
      -1 & 0
    \end{matrix}\right]
    \text{and }
    A(2) = \left[\begin{matrix}
      2 & -1 \\
      5 & -2
    \end{matrix}\right]
  \end{align*}
  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
  \begin{align*}
    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
      2 & -1 \\
      -1 & 0
    \end{matrix}\right] ,
  \end{align*}
  \jss{ Or should it be: \begin{align*}
    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
      1 & 0 \\
      2 & -1
    \end{matrix}\right] ,
  \end{align*} }
  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$. The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx. $\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity. The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.

  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
\end{example}

\begin{definition}[Biological Species Concept]
  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
\end{definition}

\begin{figure}
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    %\path [->] (A) edge[bend left] node {$c$} (B);
    \path [->] (B) edge node {\tiny $\gamma$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
    \path [->] (B) edge[loop right] node {\tiny $\lambda_{2}$} (B);
\end{scope}

\end{tikzpicture}
    \caption{Gene Regulatory Network $A(0)$.}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $(\lambda_{2} - \lambda_{1} - \frac{\gamma}{2})$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{\gamma}{2}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
    \path [->] (B) edge[loop right] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
\end{scope}

@@

\jss{Maybe we should discuss the Hartman-Grobman Theorem?}

@@

\section{Estimates of network rewiring and developmenal systems drift.} \label{sec:param}

\jss{Parameter estimates for DSD $\downarrow$}

What parameters (and other relevant factors) do we need from the literature and experiment? 
  \begin{itemize}
    \item mutation rate
    \item Transcription factor binding site appearence and loss
    \begin{itemize}
      \item In \citep{lynch2007evolution} TF binding site gain ($u_{g}$) and loss ($u_{l}$) is estimated to be,
        \begin{align*}
          u_{g} &= \frac{(L-n+1)nu}{4^{n}} \\
          u_{l} &= nu
        \end{align*}
      Where $L$ is the length of the promoter, $n$ is the number of nucleotides in a TFBS, and $u$ is the mutation rate per base pair per generation. 
    \item \citet{lynch2007evolution} further states that in prokaryotes $L$ will usually be $\leq 100$ and in eukaryotes $L$ can range from $10^{3}$ to $10^{6}$. He also cites \citep{harbison2004transcriptional} that TFBSs range from $5-10$bps in length.
    \item in \citep{lynch2014evolutionary}, they estimate
        \begin{align*}
          \tilde{P}(m) = C \left[ 3^{\ell - m}  \left( \begin{array}{c} \ell \\ m \end{array} \right) \right] e^{2N_{e}s(m)}  ,
        \end{align*}
        where $\tilde{P}(m)$ is the stationary distribution of a TFBS with match (and strength) $m$. $C$ is the normalizer constant, $m$ is the number of matches, $\ell$ is the length of the TFBS motif. The term on the right is the ratio of fixation probabilities. 
        Or in their own words, there are two interpretations: 
        \begin{quote} ``(i) For a single TFBS, it represents the long-term proportion of evolutionary time spent in the various matching states and (ii) for a set of different TFBSs under the same selective constraints, it represents the expected distribution of states at any point in time.'' \end{quote}
      \item \citet{tuugrul2015dynamics, anderson2015intermolecular} also may be helpful.
    \end{itemize}
    \item rate of both \emph{trans-} and \emph{cis-} regulatory evolution. (see Sergey's papers).
      i.e. \citep{nuzhdin2004common, fear2016buffering, graze2009regulatory, graze2012allelic, genissel2008cis}
    \item The PWN-score landscape (does binding strength increase linearly, exponentially, have a threshold, etc.) with respect to SNPs.
    \item Others? 
  \end{itemize}

\section{Adaptation}

Suppose a population is perfectly adapted to its environment -- it expresses a metabolic enzyme in response to environmental sugar availability at a rate that balances the energetic costs and benefits of the enzyme's synthesis and utility. Then, the population suddenly finds itself in a novel environment and exhibits suboptimal expression rates. How does the population adapt? Are some network mechanisms mutationally closer or farther from the desired new network connectivity?

Let $A_{\alpha}(p)$ be the $p$th system realization perfectly adapted to the first environment ($\alpha$), and let $A_{\beta}(q)$ be the $q$th system realization perfectly adapted to the second environment ($\beta$). 
\begin{align*}
  \{p,q : \min \sum_{i,j}^{n} | A_{\alpha}(p)_{i,j} - A_{\beta}(q)_{i,j} | \}
\end{align*}
In an asexual population, if there is variation in network topology, the $A_{\alpha}$ network closest to the organization of a $A_{\beta}$ network might be the fastest to adapt and outcompete all of its neighbors.

\section{Mutational Robustness}

Are any network realizations $A(p)$ more mutationally robust than others? If $s_{i,j}$ is the fitness cost of mutating element $A_{i,j}$ and $\mathbb{E}_{p}(s) = \frac{1}{n}\displaystyle\sum_{i,j}^{n} s_{i,j}$ is the mutationaly susceptibility of realization $p$, is there a $p$ such that $\mathbb{E}(s)$ is minimal? Will the population find this network orgnization?

\section{Network Ratchet}

If a network evolves to include more molecules (such as transcription factors) and thus becomes higher dimensional, can the network also remove these originally superfluous interactions or will removing these typically lead to catastrophic network dysfunction after some amount of evolutionary time and system drift has occured?

\section{Network Minimality versus Rube Goldbergness}

Are there evolutionary differences between realizing a transfer function minimally (with a reachable and observable system) or non-minimally (a system with unreachable and/or unobservable subspaces)? Can reducible systems be more or less mutationally robust? Are minimal systems energetically more efficient? Are reducible systems more or less evolvable?

\section{(Old Notes) What biological question are we asking?}

@@

\section{(Old Notes) How can we mathematically model this question?}

@@

\section{(Old Notes) List of papers with some notes.}

@@

\section{Including Deleterious Mutations and a Fitness Function}\label{sec:del}

  \begin{minipage}{0.5\textwidth}
  At the moment we focus strictly on the case where the minimal realization is in $2$ dimensions and both of its eigenvalues are real and negative (so a special case of the system from Example \ref{ex:2x2}). \jss{(Note: this is not to say the ideas below will not be generalizable).}

    The system of interest has the form,
  \begin{align*}
    \Sigma := \left\{ A = \begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \end{bmatrix}, B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, C = \begin{bmatrix} 1 & 0 \end{bmatrix} \right\}
  \end{align*}
  \end{minipage}%
  \begin{minipage}{0.5\textwidth}
    \centering
    \includegraphics[width=0.95\textwidth]{figure_2x2mutation01}%
      \captionof{figure}{The transfer function for $A = \left[ \protect\begin{array}{cc} -3 & 1 \\ 0 & -1 \protect\end{array} \right]$ and all eight $\mu(\Sigma, A_{ij})$ mutations, where $\upsilon = \pm 0.5$}\label{fig:mut}%
  \end{minipage}

Define the fitness function to be,

\begin{align}\label{eq:fitness}
  \Phi(\bar{\Sigma}) := \left\{ \begin{array}{ll}   e^{- \left| H(0) - \tilde{H}(0) \right| = - \left| \frac{k}{\lambda_{1} \lambda_{2}} - \frac{\tilde{k}}{\tilde{\lambda}_{1} \tilde{\lambda}_{2}} \right|}    &\text{if } \lambda_{1,2} < 0 \text{ and } \in \mathbb{R}, \ \det(A) \neq 0 \\
  0   &\text{if } \lambda_{1,2} \geq 0 \ \text{ or } \ \notin \mathbb{R}, \ \det(A) = 0 \end{array} \right.
\end{align}
and let $\upsilon$ be the magnitude and direction (sign) of a mutation on regulatory interaction strength. The fitness function (\ref{eq:fitness}) may me appropriate if mutations only slightly alter the dynamics of the system, as in Figure \ref{fig:mut}.

%Further let $\mu(\Sigma, A_{ij})$ be the resulting gene regulatory network following a mutation of matrix element $A_{ij}$ to system $\Sigma$ (which contains matrix $A$).
%
%  \begin{align*}
%    \Phi(\mu(\Sigma, A_{ij} \upsilon)) = \exp\left(- \left| \frac{A_{22}-A_{12}}{\lambda_{1} \lambda_{2}} - \frac{k + \upsilon \pda (A_{22}-A_{12})}{\lambda_{1} \lambda_{2} + \text{adj}(A)^{T}_{ij} \upsilon} \right|\right)   .
%  \end{align*}
%For instance, the fitness of an organism with a change of one of its entries $A_{ij}$ (by $\upsilon$) is listed,
%  \begin{align*}
%    \Phi(\mu(\Sigma, A_{11})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) + \upsilon A_{22}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{12})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} - \upsilon}{\det(A) - \upsilon A_{21}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{21})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12}}{\det(A) - \upsilon A_{12}} \right| \right) \\
%    \Phi(\mu(\Sigma, A_{22})) &= \exp \left(- \left| \frac{A_{22}-A_{12}}{\det(A)} - \frac{A_{22}-A_{12} + \upsilon}{\det(A) + \upsilon A_{11}} \right| \right)
%  \end{align*}
%
 The transfer function at $z=0$ for any GRN with mutations is, 
  \begin{align*}
    H(0, \begin{bmatrix} A_{11} + \upsilon & A_{12} + \varepsilon \\ A_{21} + \delta & A_{22} + \omega \end{bmatrix}) = \frac{A_{22} - A_{12} + \omega - \varepsilon}{ \det(A) + \omega A_{11} - \delta A_{12} - \varepsilon A_{21} + \upsilon A_{22} + \omega \upsilon - \varepsilon \delta }
  \end{align*}

>>>>

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepackage{caption, subcaption}

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

@@

%$\unobservable$

@@

% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway? 
% 
\title{Mathematical Gene Network Evolution}
\maketitle

\begin{abstract}
Abstract.
\end{abstract}

\section{Background}

@@

It is commonly taught that an organism's genome contains the heritable material 
that natural selection filters and that an organism's phenotype directly determines its
evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
undrstood molecular machinery -- and it is a major goal common to many different disciplines
within the life sciences to elucidate its form, function, and evolution. These aims are
delicately intertwined and a comphrehensive understanding of a system's evolution requires
an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system (or a system of differential equations).
Previously it has been argued that any idealized study of evolution is incomplete without
a mathematically sufficient description of the genotype, phenotype, and transformation from
one to the other. Movement in this direction is ongoing, as researhers have begun to study 
the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics.''
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in closely related species can be divergent at the molecular level.

@@

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur).

@@

Organisms often have to respond to their environments,
and the way this happens is that external input triggers a cascade of molecular signals
whose eventual result is a reaction,
whose approprateness for the situation determines fitness.
\plr{A similar situation occurs in development (...?).}
However, there are many ways to skin a cat:
especially given a number of possible molecular intermediaries,
there may be a large number of ways to construct such a molecular signalling system
that has precisely the \emph{same} input-output relationships,
and hence the same function.
Therefore, any mutation that changes one such system to another, equivalent version
will be neutral,
and may well become polymorphic in a population if there are no ill effects in heterozygotes.
Certainly, mutations that move a system away from optimal functioning will also appear,
but in very large populations will not rise to high frequency.
\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}

@@

\begin{example}[Example: Oscillator.] \label{ex:osc}
Motivate the system:
\begin{align*}
    A &= \left[\begin{matrix} 
        0 & 1 \\ 
       -1 & 0 
    \end{matrix}\right]
\end{align*}
as a special case of the below.

@@

\begin{example}[Hyrbid Inviability]
  Let, \begin{align*}
    A(0) =\left[\begin{matrix}
      0 & 1 \\
      -1 & 0
    \end{matrix}\right]
    \text{and }
    A(2) = \left[\begin{matrix}
      2 & -1 \\
      5 & -2
    \end{matrix}\right]
  \end{align*}
  from Example \ref{ex:osc} ($B=[1 ,1]^{T}$ and $C=[1 0]$). 
If hybrids are formed by exchanining genes (swapping rows), they hybrid,
  \begin{align*}
    \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
      2 & -1 \\
      -1 & 0
    \end{matrix}\right] ,
  \end{align*}
  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$. The time dynamcs of molecular species $1$ in the hybrid is $y' =\mathrm{e}^{t}\, \cosh\!\left(\sqrt{2}\, t\right)$, with species $1$ quickly heading towards infinity. The dynamics of species $1$ in the parental networks is, $y = \sin(t) + \cos(t)$, where species $1$ continuously oscillates.
\end{example}

\begin{figure}
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (0,-4) {$x_{2}$};
    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    %\path [->] (A) edge[bend left] node {$c$} (B);
    \path [->] (B) edge node {\tiny $\gamma$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
    \path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
\end{scope}

\end{tikzpicture}
    \caption{Gene Regulatory Network $A(0)$.}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (0,-4) {$x_{2}$};
    \node[shape=rectangle, dashed] (U) at (3,-2) {$u(t)$};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $(\lambda_{2} - \lambda_{1} - \frac{\gamma}{2})$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{\gamma}{2}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
    \path [->] (B) edge[loop left] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
\end{scope}

@@

@@

\section{What biological question are we asking?}

@@

\section{How can we mathematically model this question?}

@@

\section{List of papers with some notes.}

@@

\section{Notes on what to do when allowing deleterious mutations}

Epistasis, coming from mutations deleterious in some networks but not in others
(come from incompatibilities).

Describe heterosis.
